---
title: "phylo101 Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phylo101 Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# phylo101: Introduction to Phylogenetics

## Load required packages

```{r setup}
library(adegenet)
library(ape)
library(ggtree)
library(ggimage)
library(readr)
library(phylo101)
```

## Check that metadata is loaded

```{r}
d <- USFluMetadata
print(d)
```

## Check that sequences are loaded

```{r}
s <- USFluSeqs
print(s)
```

# Models of DNA Evolution

## Calculate pairwise distances

Wikipedia entry about models: <https://en.wikipedia.org/wiki/Models_of_DNA_evolution>

I would run a few different models, starting with "raw". Definitely include "K80" (see below) and then pick 1-2 other models to run. Briefly explain why one model might be better than another (hint: the example data are seasonal flu sequences). ?dist.dna() can be helpful

You can visualize the pairwise distances as a distance matrix. This isn't a super practical figure that you would include in most main manuscripts, but is useful for showing that the distance matrix exists and helps to show what a distance matrix is.

```{r}
# raw
```

```{r}
# Kimura, 1980
dist <- dist.dna(s, model = "K80")
```

# Methods of Phylogenetic Tree Construction

## Construct phylogenetic tree and create tree object

Construct trees using one of the distance matrices created above, preferably the "best" one (if available). Then run a few methods of constructing the phylogeny. I've provided some suggestions- but feel free to use different methods. After each model I would create a basic tree visualization (using ggtree makes the most sense here).

```{r}
# Neighbor-Joining 2.0
t1 <- bionj(dist)
```

```{r}
# Neighbor-Joining
t2 <- nj(dist)
```

```{r}
# UPGMA
## I suggest adding the phangorn package for this

```

```{r}
# Minimum Evolution

```

# Phylogenetic Tree Visualization

## Plot/visualize tree using ggtree

```{r}
# open tree in new window for optimal viewing
ggtree(t1) +
  geom_tiplab()
```

## Add metadata to the tree

```{r}
p <- ggtree(t1) %<+% d
```

## Annotate tree tip labels using color

### Color by year

```{r}
library(viridis)
p + geom_tiplab(aes(color = year), size = 3) +
    scale_color_viridis_c() +
    theme_tree2()
```

### Color by state

```{r}
p + geom_tiplab(aes(color = state), size = 3) +
    scale_color_hue(n = 15) +
    theme_tree2()
```
